prolfqua - An R-package for Analysing Proteomics Label Free Quantification Experiments

### 2022-04-01 inst/MyArticle/paper.Rmd
update bib

Label-Free Quantification (LFQ) Proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs are implemented with increasing numbers of samples, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a factor or interactions are significant, or differences between factor levels can be estimated, and their significance tested. Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [@subramanian2005gene] or validation experiments.

@@

When analyzing parallel group designs using a single explanatory variable, we can use all packages for the statistical data analysis of the data; however, if we want to model factorial designs or repeated measurements, we can use only some of them. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages.

@@

We were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables to call various machine learning (ML) methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, all of them fit some type of linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, some of them apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented (OO) design, with R linear models as a core, where features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")` or variance shrinkage could be easily enabled (see Figure \@ref(fig:ContrastUML)).

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the fold change was imputed in such a way are explicitly annotated in the result table.

Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020].

@@

@@

@@

[TODO(wew): Goes in the discussion or introduction, and give references for normalization e.g. VSN paper.] However, an essential parameter, not discussed further, is the normalization of the intensities within the samples to make them comparable. For this benchmark dataset, the fraction of regulated proteins (_E. coli_) is relatively large. Therefore, to calibrate the measurement, we used human proteins only. In practice, it is unknown which proteins do and which don't change. Therefore, the assumption used when normalizing, scaling the data, and thus calibrating the measurement is that only a small subset of the proteins in the dataset is differentially regulated.

[TODO(wew): Goes to discussion] The method used to infer protein intensities from more peptide intensities is an additional important factor. The original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. When using those, _proDA_ performs significantly worse than _proDA_ when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_{10}(\textrm{t-statistics}) = 66%$ and $pAUC_{10}(\textrm{t-statistics}) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details).

@@

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, for instance, to perform gene enrichment analysis, the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference further. However, the observed change is minuscule.

>>>>

Protein quantification proteomics Experiments enable monitoring relative abundances of thousands of proteins in biological samples. For most studies, parallel group designs are used, e.g., treatment versus control are used [@de2022apoe2, @laubscher2021baf]. More recently, more complex experimental designs with increasing numbers of samples are studid, i.e., factorial designs and longitudinal studies (time series), sometimes with repeated measurements on the same subject [@tan2022proteomic, @meier2021reduced]. The obtained data can be modeled using linear fixed-, mixed-, or random-effects models [@Bates2015FittingJSS]. Based on these models, tests can be applied to examine whether a specific factors and their interactions are significant. Furthermore it can be tested if differences between factor levels can be estimated, and their significance tested. Proteins can be ordered or grouped using the estimated fold-change, $t$-statistics, scaled $p$-value, or false discovery rate (FDR) and then subjected to gene set enrichment or over-representation analysis [@subramanian2005gene] or validation experiments.

@@

When analyzing parallel-group designs using a single explanatory variable, we can use all R packages; however, we can use only some of them if we want to model factorial designs or repeated measurements. Table \@ref(tab:tableOverview) gives an overview of the models and features supported by these packages. We see that packages such as $limma$ and $proDA$ allow us to a comprehensive variety of hypotheses; however, in-depth knowledge of design matrix specification is required [@law2020guide].

@@

We were inspired by the R package `r BiocStyle::CRANpkg("caret")` [@caret2008], which enables to call various machine learning (ML) methods, making selecting the best ML algorithm for your problem easy. We aimed for a similar R package for the analysis of LFQ data. However, after examining the R packages for modeling proteomics quantification data, we observed that model specification, input, and output formats differ widely and wildly. However, many of these packages fit linear model either to peptide or protein intensities, determine contrasts among conditions, and afterward, apply empirical Bayes variance shrinkage. Therefore, the revised goal was to provide a modular object oriented (OO) design, with R linear models as a core, where features such as $p$-value aggregation, e.g., `r BiocStyle::Biocpkg("PECA")` or variance shrinkage could be easily enabled (see Figure \@ref(fig:ContrastUML)).

An important aspect of mass spectrometric data are missing values. Missing values lead to groups of varying size. Linear and mixed effect models can handle unbalanced designs, i.e. different group sizes. As long as there is at least one observation per group they do produce unbiased estimates. We use this feature instead of imputing missing observations. 
However, if there are no observation in one of the groups, then contrasts involving this group can not be estimated. To address this, for the group without observations, we use a point prior equal to the average of $3\%$ smallest group intensities means in the dataset and infer the variance using the pooled the variance of other groups. Proteins for which the fold change was imputed in such a way are explicitly annotated in the result table. 
Other means of modelling missing information in proteomics data using the Hurdle model is discussed by [@goeminne2020msqrob], while the R package `r BiocStyle::Biocpkg("proDA")` models missingness using probabilistic dropout models [@bioxrvproDA2020]. 
We show using a benchmark dataset that this, although trivial approach, can outperform methods based on data imputation or dropout models.

@@

An essential parameter of any protein quantification pipeline, not discussed further, is the normalization of the intensities within the samples to make them comparable [@pursiheimo2015optimization]. All benchmarks we run with the same data normalization method. The method used to infer protein intensities from peptide intensities is an additional important factor. For instance, the original _proDA_ publication uses MaxLFQ [@cox2014accurate] protein estimates. However, when using those protein abundance estimates, _proDA_ performs significantly worse when using the protein estimates obtained from peptide intensity estimates using Tukey's median polish,  $pAUC_{10}(\textrm{t-statistics}) = 66%$ and $pAUC_{10}(\textrm{t-statistics}) = 73%$ respectively (see Articles on [@prolfquaGithubIO] for more details). Last but not least, the software, identifying proteins, and generating the quantification values can significantly contribute to the performance of the entire pipeline [@Cox2008MaxQuant,@yu2020fast]. However, here we focus on the comparison of the various modeling approaches.

@@

For this benchmark dataset, the fraction of regulated proteins (_E. coli_) is relatively large. Therefore, to calibrate the measurement, we used human proteins only. In practice, it is unknown which proteins do and which don't change. Therefore, the assumption used when normalizing, scaling the data, and thus calibrating the measurement is that only a small subset of the proteins in the dataset is differentially regulated.

@@

@@

We can conclude that if we want to order the proteins according to the likelihood of being differentially regulated, for instance, to perform gene set enrichment analysis, the $t$-statistic is better suited than the fold-change estimate. Modeling the degrees of freedom when computing the $p$-values might improve the inference further. However, the observed change is minuscule.
